---
title: "Implementation - stationary use"
author: "Juan S Vargas"
date: "2023-06-16"
output: html_document
---

# Implementation for the moveSTIR paper

Here I will use simulated tracks following the approach by Scharf et el. [@Scharf2018]. This is similar to the method used in the examples in the original moveSTIR paper, but it has an additional social smoothing kernel which allows to account for animals moving closer to each other, or moving together. I will use the simple, independent tracks first, and then make changes to test different case studies.


## Independent movement

First we will test the framework for a simulated scenario where movement is independent across individuals. While there could be overlap in the resulting utilization distributions and points where the tracks cross each other, there should be no significant temporal correlation in the location histories.
``` {r import-plot-tracks}
library(sp)
library(adehabitatHR)
library(tidyverse)
library(cowplot)
```

```{r import plot indep tracks}
sim_tracks <- lapply(list.files("../data", "indep(.*).csv",full.names = TRUE), read.csv)
# join together
sim_tracks_db <- do.call(rbind, sim_tracks)[,2:3] |> cbind(ind = rep(seq_along(sim_tracks), sapply(sim_tracks,nrow))) 
# Create a figure to illustrate the example. This figure will have the tracks of multiple individuals, then their respective UDs, as well as the UD product.
ggplot(sim_tracks_db)+geom_path(aes(x,y,color = factor(ind)))+
  theme_minimal()+
  coord_sf()+
  labs(color = "Individual")
# plot(sim_tracks_db$x, sim_tracks_db$y, type = "n", asp = 1, axes = F, ann = F)
# for (i in unique(sim_tracks_db$ind)) {
#   lines(sim_tracks_db[sim_tracks_db$ind == i, ], col = hcl.colors(length(sim_tracks), palette = "Set 2")[i], lwd  = 1.2)
# }
# legend(x = "left",legend = 1:length(sim_tracks), col = hcl.colors(length(sim_tracks),palette = "Set 2"),border = F, lty = 1)
```
Let's say that for this simulation the units are kilometers, and a reasonable contact distance for parasite transmission is 10 meters.



We then get the utilization distribution for every track. In this case I am using simple kernel density estimation, i.e. I'm not using any mechanistic movement model, but the same procedure would apply.

```{r get-plot-kernels, fig.height=3}
# create the spatialPoints object with id column
sim_tracks_sppoints <- SpatialPointsDataFrame(sim_tracks_db[,1:2], data = data.frame(id = sim_tracks_db$ind))
# get the UDs with a common grid
kuds <- kernelUD(sim_tracks_sppoints, same4all = TRUE)

# Get the cell area, for FOI calculation
cell_area <- kuds[[1]]@grid@cellsize |> prod()

sapply(kuds, plot)
```
We multiply these UDs pairwise to get the probability $p_i(x)p_j(x)$. This is the first component of the equation. The figure below shows an example of the result; we get a matrix of the same size, but the highest values will be the areas of overlap between individuals.

```{r UD-prod, fig.height=3}
# Calculate the per-cell product of expected values
udprod <- function(k1, k2) {k1@data*k2@data}

# Plot one example
plot(SpatialPixelsDataFrame(coordinates(kuds[[1]]), 
                            data = data.frame(udprod(kuds[[1]],kuds[[2]]))))
# Get all the products in a list
ind_combs <- expand.grid(ind2 = seq_along(kuds), ind1 = seq_along(kuds))
ind_combs <- ind_combs[ind_combs$ind1!=ind_combs$ind2,]
UDprods <- apply(ind_combs, 1, \(x) udprod(kuds[[x[2]]],kuds[[x[1]]]), simplify = FALSE)
```
For the remaining components, i.e. the standard deviation and the correlation, we need the location history for every individual. The location history is a $T$ by $X$ binary matrix, with $T$ the number of discrete time steps and $X$ the number of sites, that shows in which cell the individual was at every time point. We do this finding which grid cell center is closest for every spatial point.

```{r loc-hist}
# Next we need to determine the location histories. The location history is a matrix with time in rows, locations (grid cells) in columns. Matrix values represent whether the individual was (1) or was not (0) at that location at that time. So, we need to assign a cell to every location.

# Get the distance between the grid cell centers and every point in the tracking history. This creates a numeric matrix with points in rows, and grid cell centers in columns. 
point_grid_dists <- spDists(sim_tracks_sppoints, kuds[[1]])

# Create a binary matrix and split it by individual
loc_histories <-  apply(point_grid_dists, MARGIN = 1, \(x) x==min(x)) |> t() |>
  split(factor(sim_tracks_db$ind)) |> 
  lapply(matrix, ncol = ncol(point_grid_dists))

#### Important, the grid cells are not in order!!!

# Create a function to do this more formally (In prgress)
# get_lochist <- function(s, k) {
#   pt_grid_dist <- spDists(s,k)
#   l <- apply(pt_grid_dist, 1, \(x) x == min(x))
#   return(l)
# }
```
The location histories are iteratively resampled, to select times in the history of $i$ and match them with times in the history of $j$ $s$ time units before. Once we have these subsets of the location histories, we calculate the covariance across matching matrix columns for two different individuals. Another way is to divide the covariance, expressing it as the product of the per-cell standard deviation as $\sigma=\sqrt{\hat p(1-\hat p)}$, where $\hat p$ is the estimated probability for a binomial process, and the correlation across the same cells for two individuals. It would seem there is no gain in speed in calculating the correlation as opposed to the covariance.

```{r covariance}
# We need to integrate the product of the covariance between the presence of i at x at time t, and the presence of j at x at time u = t-s, across all lags s. This is then multiplied by e^(-mu*s)
get_cov <- function(lh) {
  # matrix of combinations. A t*t matrix of difference in time. This is used to
  # find the row-col combinations corresponding to a given lag
  lag_mat <- round(outer(sim_tracks[[1]]$time, sim_tracks[[1]]$time, "-"),5)
  lags <- unique(lag_mat[lower.tri(lag_mat,diag = TRUE)])
  lags <- lags[-length(lags)] # remove final value that would have only one possible combination
  
  # For every lag s, find the rows in host 1 and host 2 corresponding to times t
  # and t-s, calculate the covariance
  cell_covs <- replicate(n = nrow(ind_combs), 
                         expr = matrix(0, nrow = length(lags), 
                                       ncol = ncol(loc_histories[[1]])),
                         simplify = FALSE)
  
  for (r in 1:nrow(ind_combs)) {
    i <- ind_combs[r,2]
    j <- ind_combs[r,1]
    covmat <- cell_covs[[r]]
    for (s in seq_along(lags)) {
      l <- lags[s]
      # find position where those lags occur
      indices <- which(lag_mat==l, arr.ind = T)
      # subset host position matrices with indices
      h1 <- lh[[i]][indices[,1],]
      h2 <- lh[[j]][indices[,2],]
      ovlpcells <- which((colSums(h1)*colSums(h2))>0)
      if(length(ovlpcells)==0) next
      if (length(ovlpcells)>0) {
        h1sub <- h1[,ovlpcells]
        h2sub <- h2[,ovlpcells]
        covmat[s,ovlpcells] <- diag(as.matrix(cov(h1sub,h2sub)))# only the diagonal as we care only about same cells
      }
    }
    cell_covs[[r]] <- covmat
  }
  return(cell_covs)
}

cell_covs <- get_cov(loc_histories)
# multiply by the decay, and sum across 
cell_cov_int <- lapply(cell_covs, "*", exp(-nu*lags)) |> lapply(colSums)
```
We obtain a matrix with cells in rows, and combinations of individuals in columns. For every column we need to add the product of the UDs at that cell, and then scale by the epidemiological parameters to obtain the cell FOI.

```{r foi}
calculate_cell_foi <- function(ud, ci) {
  # ud are the products of the utilization distributions, cv is the list of integrated cell-specific covariance multiplied by the exponential decay
  UDmu <- lapply(ud, "*", 1/nu)
  compsum <- mapply("+", ci, UDmu,USE.NAMES = F) # add the UD product and the integrated covariance term
  cell_foi <- sapply(compsum, "*", beta*lam) # multiply by the epidemiological parameters
  cell_foi <- replace(cell_foi, cell_foi<0, 0)  # substitute negative values for 0s
  return(cell_foi)
}

cell_foi_indep <- calculate_cell_foi(UDprods, cell_cov_int)
```

We obtain a matrix where each column is a combination of two individuals, and every row is a cell in the landscape. The matrix values are the estimated FOI experienced by individual 1 from individual 2. By adding across the columns we can estimate the total FOI for that pair of individuals. From this matrix we can also calculate the total FOI that an individual experiences from all other individuals, either at a specific location or across the whole landscape.

```{r tracks-gg}
sim_tracks_db %>% ggplot()+geom_path(aes(x,y,color = factor(ind)), size=1)+theme_void(base_size = 14)+coord_sf()+labs(color="Individual")+theme(legend.position = "left") -> gg.tracks
```
```{r pairwise-foi}
# total FOI from for every pair of individuals
data.frame(ind_combs, foi = colSums(cell_foi_indep)) |>
  ggplot()+
  geom_raster(aes(factor(ind1),factor(ind2, levels = 6:1),fill=foi))+
  theme_minimal(base_size = 14)+
  labs(fill = "FOI", y = "Acquiring individual", x = "Depositing individual")+
  scale_x_discrete(breaks = 1:6, position = "top",expand = c(0,0))+
  scale_y_discrete(breaks = 1:6, expand = c(0,0))+
  theme(aspect.ratio = 1, panel.grid = element_blank()) -> gg.pairfoi

cowplot::plot_grid(gg.tracks,gg.pairfoi)
```

We can see that individual 1 has hardly any overlap, and its FOI values are the lowest. The product of the UDs is symmetrical, but the difference in order of arrival at a given site creates differences due to the covariance term.

```{r ind-foi}
# total FOI experienced across from all hosts combined
tapply(colSums(cell_foi_indep), ind_combs$ind1, sum)
# total FOI contributed to all other hosts
tapply(colSums(cell_foi_indep), ind_combs$ind2, sum)
```
```{r plot-cellfoi}
# Plot one example
plot(SpatialPixelsDataFrame(coordinates(kuds[[1]]), 
                            data = data.frame(rowSums(cell_foi_indep))))
```
```{r st-dev}
# # We can calculate the sd from the estimated p_i. Can we?
# stdevs_est <- lapply(kuds, \(x) sqrt(as.matrix(x)*(1-as.matrix(x))))

# calculate standard deviation from the data. For a Bernoulli process, the sd is sqrt(p(1-p)), where p is the column mean for the binary location history matrix.
getcellSD <- function(h) sqrt(colMeans(h)*(1-colMeans(h)))
# apply(h, 2, \(x) sqrt(mean(x)*(1-mean(x))))
stdevs <- lapply(loc_histories, getcellSD)
```
```{r}
sapply(cell_covs, rowMeans) %>% as_tibble() %>% add_column(lag = lags) %>% 
  pivot_longer(cols = starts_with("v")) %>% 
  ggplot()+geom_point(aes(lag,value,color = name))
```

## Joint movement

Now we will do the same for a case where there is joint movement for two hosts, but the rest are moving independently.

```{r joint-mov}
## 1. Import data, calculate UDs
sim_tracks <- lapply(list.files("../data", "joint(.*).csv",full.names = TRUE), read.csv)
sim_tracks_db <- do.call(rbind, sim_tracks)[,2:3] |> cbind(ind = rep(seq_along(sim_tracks), sapply(sim_tracks,nrow))) 
# create the spatialPoints object with id column
sim_tracks_sppoints <- SpatialPointsDataFrame(coords = sim_tracks_db[,1:2], 
                                              data = data.frame(id = sim_tracks_db$ind))
# get the UDs with a common grid
kuds <- kernelUD(sim_tracks_sppoints, same4all = TRUE)
# Get the cell area, for FOI calculation
cell_area <- kuds[[1]]@grid@cellsize |> prod()

## 2. UD product
ind_combs <- expand.grid(ind2 = seq_along(kuds), ind1 = seq_along(kuds))
ind_combs <- ind_combs[ind_combs$ind1!=ind_combs$ind2,]
UDprods <- apply(ind_combs, 1, \(x) udprod(kuds[[x[2]]],kuds[[x[1]]]), simplify = FALSE)

## 3. Location history
# Get the distance between the grid cell centers and every point in the tracking history. 
point_grid_dists <- spDists(sim_tracks_sppoints, kuds[[1]])
# Create a binary matrix and split it by individual
loc_histories <-  apply(point_grid_dists, MARGIN = 1, \(x) x==min(x)) |> t() |>
  split(factor(sim_tracks_db$ind)) |> 
  lapply(matrix, ncol = ncol(point_grid_dists))

## 4. Get the covariance
cell_covs_joint <- get_cov(loc_histories)

## 5. multiply by the decay, and sum across to integrate
cell_cov_int <- lapply(cell_covs_joint, "*", exp(-nu*lags)) |> lapply(colSums)

## 6. Combine the integrated covariance term and the  UD product, and scale by the epi parameters
cell_foi_joint <- calculate_cell_foi(UDprods, cell_cov_int)
```

```{r plot_joint_sim}
## 7. plots
data.frame(ind_combs, foi = colSums(cell_foi_joint))|>
  ggplot()+
  geom_raster(aes(factor(ind1),factor(ind2, levels = 6:1),fill=foi))+
  theme_minimal(base_size = 14)+
  labs(fill = "FOI", y = "Acquiring individual", x = "Depositing individual")+
  scale_x_discrete(breaks = 1:6, position = "top",expand = c(0,0))+
  scale_y_discrete(breaks = 1:6, expand = c(0,0))+
  theme(aspect.ratio = 1, panel.grid = element_blank()) -> gg.pairfoi

data.frame(ind_combs, foi = beta*lam*sapply(UDprods, sum)) |>
  ggplot()+
  geom_raster(aes(factor(ind1),factor(ind2, levels = 6:1),fill=foi))+
  theme_minimal(base_size = 14)+
  labs(fill = "FOI", y = "Acquiring individual", x = "Depositing individual")+
  scale_x_discrete(breaks = 1:6, position = "top",expand = c(0,0))+
  scale_y_discrete(breaks = 1:6, expand = c(0,0))+
  theme(aspect.ratio = 1, panel.grid = element_blank())-> gg.pairfoiindep

gg.tracks <- sim_tracks_db %>% ggplot()+geom_path(aes(x,y,color = factor(ind)), size=1)+theme_void(base_size = 14)+coord_sf()+labs(color="Individual")+theme(legend.position = "left")

plot_grid(gg.tracks,plot_grid(gg.pairfoi,gg.pairfoiindep,nrow=2, align = "v"))
```

There is a clear difference in FOI between the two individuals who are moving together, and the rest. The estimated FOI is at least twice as high for that pair as for the others. This difference is not due solely to the degree of overlap; while they do have the highest degree of overlap, they also overlap similarly with other individuals,

What happens if we do not incorporate the covariance term? This is, what if we assumed independent movement, and only used the product of the UDs, how much error would we be incurring?

```{r, fig.height=3}
tibble(ind_combs, foi_full = colSums(cell_foi_joint), foi_ud = beta*lam*sapply(UDprods,sum)) %>% mutate(udprop = (foi_full-foi_ud)/foi_full) %>% 
  ggplot()+geom_raster(aes(ind1,ind2, fill = udprop))+
  # scale_fill_gradient2(midpoint = 1)+
  theme_minimal()+
  theme(aspect.ratio = 1, panel.grid = element_blank())+
  labs(x = "Depositing individual", y = "Acquiring individual", fill = "Cov contribution")
```

In cases where there is little overlap the product of the UDs makes up the majority of the FOI. However when the individual tracks do overlap, the covariance term seemingly makes up the majority of the FOI.

Let's see how the covariance varies for the different combinations of individuals.

```{r}
# First let's see how the covariance term changes in response to s, the lag
sapply(cell_covs_joint, rowMeans) %>% as_tibble() %>% add_column(lag = lags) %>% 
  pivot_longer(cols = starts_with("v")) %>% 
  ggplot()+geom_point(aes(lag,value,color = name))+
  theme_classic(base_size = 14)+
  labs(x = "Lag s", y = "Covariance")
```

We see there is some periodic behavior with respect to the lag $s$ for a couple of the covariance series, the two combinations of the individuals that are moving together. The covariance is positive and high for small lags but decreases sharply. For the other series—with independent movement—there is no clear trend.

Let's take a look at the covariance term and how it's related to overlap.

```{r}
m1 <- kerneloverlap(sim_tracks_sppoints)
m2 <- diag(6)
m2[as.numeric(rownames(ind_combs))] <- sapply(cell_cov_int, sum)
plot(m1,m2)
```

```{r}
sapply(cell_covs_joint, rowMeans) %>% as_tibble() %>% add_column(lag = lags) %>% 
  pivot_longer(cols = starts_with("v")) %>% 
  ggplot()+geom_boxplot(aes(name,value))+
  theme_classic(base_size = 14)+
  labs(x = "Pair", y = "Covariance")
```

